Estimating the distribution of block sizes as a function of average transaction rates

### 2015-12-04 paper/paper.tex
Fixed some typos

For the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate. This is not surprising since Bitcoin usage has gone up over time. A block is around 10 minutes on average, so 5000 blocks takes a little over a month on average. It is possible that on a short enough time-scale, transaction rates are approximately constant. So we fit the gamma distribution to 1000 blocks at a time --- representing about a week --- by breaking the original 5000 blocks into five groups. For each of these groups the 95\% credible interval does not contain one --- the lowest lower bound is 1.17. Similarly for groups of 500 blocks, each representing about 3.5 days, the lowest lower bound is about 1.08. So we can safely conclude that the transaction rate is not constant. A likely reason for this is seasonality in transaction rates --- Bitcoin trades are probably more likely to happen during certain times of day than others, based on when relevant markets are open and when the bulk of Bitcoin users are awake. The {\it mean} transaction rate averaging over e.g. seasonality is probably approximately constant for time-scales this small, however. In any case, transaction rates are nonconstant even on short time-scales, which motivates modeling them as such.

@@

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official time-stamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' time-stamps, occasionally a block has a time-stamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid time-stamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the block chain, so the upper bound is not observed, but the lower bound is at our disposal.

@@

Since the observation times are observed with error, it is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form for any $tau$ and $\delta>0$. This simplifies fitting the model because it allows us to write down a relatively simple exact likelihood function, facilitating both maximum likelihood estimation and Markov chain Monte Carlo methods. The parameters of the gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the transaction data increments. We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.

@@

Combining these pieces we obtain the following model. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k-1} + \delta_k} - x_{\tau_{k-1}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
\delta_{k} = \tau_{k} - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim half-Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$. Note that the half-Caucy distribution is the Cauchy truncated to the positive real line. Table \ref{tab:modelfit} contains the results of fitting the model.

>>>>

For the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is not one and that a constant transaction rate is not appropriate. This is not surprising since Bitcoin usage has gone up over time. A block is around 10 minutes on average, so 5000 blocks takes a little over a month on average. It is possible that on a short enough time-scale, transaction rates are approximately constant. So we fit the gamma distribution to 1000 blocks at a time --- representing about a week --- by breaking the original 5000 blocks into five groups. For each of these groups the 95\% credible interval does not contain one --- the lowest lower bound is 1.17. Similarly for groups of 500 blocks, each representing about 3.5 days, the lowest lower bound is about 1.08. So we can safely conclude that the transaction rate is not constant. A likely reason for this is seasonality in transaction rates --- Bitcoin trades are probably more likely to happen during certain times of day than others, based on when relevant markets are open and when the bulk of Bitcoin users are awake. The {\it mean} transaction rate averaging over e.g. seasonality is probably approximately constant for time-scales this small, however. In any case, transaction rates are nonconstant even on short time-scales, which motivates modeling them as such.

@@

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining how transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official time-stamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' time-stamps, occasionally a block has a time-stamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid time-stamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the block chain, so the upper bound is not observed, but the lower bound is at our disposal.

@@

Since the observation times are observed with error, it is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form for any $\tau$ and $\delta>0$. This simplifies fitting the model because it allows us to write down a relatively simple exact likelihood function, facilitating both maximum likelihood estimation and Markov chain Monte Carlo methods. The parameters of the gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the transaction data increments. We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.

@@

Combining these pieces we obtain the following model. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k-1} + \delta_k} - x_{\tau_{k-1}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
\delta_{k} = \tau_{k} - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim \mathit{half-Cauchy}(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$. Note that the half-Cauchy distribution is the Cauchy truncated to the positive real line. Table \ref{tab:modelfit} contains the results of fitting the model.
